A tool set for random number generation on GPUs in R

### 2021-11-12 pkg/gpuRandom/inst/documents/paper1_2021_5/paper1.tex
update paper 1
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/diseasemapping@2960 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

%THE PROBLEM:  each process needs to be independent, there's a risk of overlap.  solution: streams.  mrg31kp3 a stream is a vector of three positive integers.  initial state is first stream, compute new streams sequentially.
\subsection{Creating streams}\label{createstreams}
Random number generation is more complicated when the RNG algorithm is run in parallel processes in \proglang{R}, there is a risk that the generated sequences of random numbers have correlation. \pkg{clrng} uses multiple distinct streams that are used in work-items that executes in parallel on a GPU device. The popular way of obtaining multiple streams is to take an RNG with a long period and cut the RNG's sequence into very long pieces, these pieces are adjacent and of equal length $Z$, so each piece can be used as a separate stream. Creating a new stream amounts to computing its starting point (seed). In general, a stream object contains three states: the current state of the stream, the initial state of the stream (or seed), and the initial state of the current substream (by default it is equal to the seed). Streams are created sequentially in the way that whenever the user creates a new stream, the software automatically jumps ahead by Z steps to find its initial state, and the three states in the stream object are set to it. Each of these streams can also be partitioned into substreams with equally-spaced starting points \citep{l2002object, rLEC15a}, which are occasionally useful.

In \proglang{clRNG} library, for the MRG31k3p RNG, the entire period length of approximately $2^{185}$ is divided into approximately $2^{51}$ non-overlapping streams of length $Z = 2^{134}$. Each stream can be further partitioned into substreams of length $W = 2^{72}$, although this is not currently implemented in \pkg{clrng}. The state (and seed) of each stream is a vector of six 31-bit integers. This size of state is appropriate for having streams running in work-items on GPU cards, while providing a sufficient period length for most applications. The initial state of the first stream (also called ``initial seed for the package'' in \proglang{clRNG} and ``initial'' in \pkg{clrng}) for the MRG31k3p is by default $(12345, 12345, 12345, 12345, 12345, 12345)$.

The function \fct{clrngMrg31k3pCreateStreams} from \proglang{clRNG} creates streams on the host using the MRG31k3p RNG. If we want to use the streams in work-items on a GPU device, the streams have to be copied from the host to the global memory of the GPU device, and then each work-item copies the current states only of the streams to its private memory. \pkg{clrng} is able to create streams both on the host and on the GPU device. \fct{createStreamsCpu} is an interface function of \fct{clrngMrg31k3pCreateStreams} that creates streams on host as \proglang{R} matrices, see the \proglang{R} output below, where we created 4 streams on the host using \fct{createStreamsCpu}, after transposing \code{myStreamsCpu}, each column of the matrix is a stream, creating streams on host makes it easy for us to view and check the streams behavior. Substreams are not created in \pkg{clrng} because currently we use one stream per work-item. 
\begin{knitrout}
\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
\begin{verbatim}
# creating streams on CPU
library("gpuR")
library("clrng")
myStreamsCpu <- createStreamsCpu(n=4, initial=12345)
t(myStreamsCpu)
##               [,1]       [,2]       [,3]       [,4]
## current.g1.1 12345  336690377  502033783  739421137
## current.g1.2 12345  597094797 1322587635 1475938232
## current.g1.3 12345 1245771585 1964121530  730262207
## current.g2.1 12345   85196284 1949818481 1630192198
## current.g2.2 12345  523477687 1607232546  324551134
## current.g2.3 12345 2094976052 1462898381  795289868
## initial.g1.1 12345  336690377  502033783  739421137
## initial.g1.2 12345  597094797 1322587635 1475938232
## initial.g1.3 12345 1245771585 1964121530  730262207
## initial.g2.1 12345   85196284 1949818481 1630192198
## initial.g2.2 12345  523477687 1607232546  324551134
## initial.g2.3 12345 2094976052 1462898381  795289868
\end{verbatim}
\end{kframe}
\end{knitrout}
\fct{createStreamsCpu} has two arguments:
\begin{itemize}
\itemsep0em 
  \item \code{n}: number of streams to create.
  \item \code{initial}: vector of length 6, recycled if shorter.
\end{itemize}
We can have streams on GPU by either converting the \code{myStreamsCpu} to a `\code{vclMatrix}' or producing streams using \fct{createStreamsGpu}. \fct{createStreamsGpu} is adapted from the clrngMrg31k3pCreateStreams() and creates streams directly on a GPU device. Setting a seed same as the MRG31k3p default seed in \fct{createStreamsGpu}, we produce on device exactly the same streams as those created on host as follows. Notice that the code is just for showing \fct{createStreamsGpu} and \fct{createStreamsCpu} are \proglang{R} interface functions for \fct{clrngMrg31k3pCreateStreams} on GPU and CPU respectively, in practice it is not recommended to create a new stream that has the same initial seed as another existing stream in one simulation program.
\begin{knitrout}
\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
\begin{verbatim}
# creating streams on GPU
myStreamsGpu1 = vclMatrix(myStreamsCpu)
myStreamsGpu2 = createStreamsGpu(n=4, initial=12345)
range(as.matrix(myStreamsGpu1 - myStreamsGpu2))
## [1] 0 0
\end{verbatim}
\end{kframe}
\end{knitrout}
\fct{createStreamsGpu} also has the two arguments \code{n} and \code{initial}. \fct{createStreamsGpu} is similar to \code{.Random.seed} in \proglang{R}. By setting the RNG status with the \proglang{R} function \fct{set.seed}, we are able to repeat the sequence of random seeds, which is an important quality criteria of RNGs. Similarly in \fct{createStreamsGpu} and \fct{createStreamsCpu}, with the \code{initial} argument, we make the sequence of streams repeatable.

\subsection{Using streams to generate uniform random numbers}
The streams created sequentially on a GPU are then used in work-items that executes in parallel on a GPU device. In \pkg{clrng} each work-item takes one distinct stream to generate random numbers, so the number of streams should always be equal to the number of total work-items in use. The main part of the kernel (\proglang{OpenCL} functions for execution on the device) for generating uniform random numbers is shown in Listing \ref{lst:uniformkernel}. Users can set \code{verbose>1} in \fct{runif} to print out the kernels. Kernels are written in the \proglang{OpenCL C} language, in which \code{__kernel} declares a function as a kernel, pointer kernel arguments must be declared with an address space qualifier, for example \code{__global} and \code{__local}. Here the pointers to \code{streams} and to the output matrix \code{out} on the global memory are passed to the kernel as arguments. \code{Nrow} and \code{Ncol} represent row and column number of the matrix \code{out} respectively. \code{NpadCol} is the internal number of columns of \code{out}, \code{NpadStreams} is the internal number of columns of \code{streams} (these variables are defined in macros not shown here). Each stream's current state is copied to the private memory of each work-item by the function \fct{streamsToPrivate}, where \code{g1} and \code{g2} point to the first three and second three elements of stream states. \code{index} represents the position of the work-item when mapping it to a one-dimensional data structure. The function \fct{clrngMrg31k3pNextState} generates an uniform random integer called \code{temp} between 1 and 2147483647,  the value of \code{temp} is then scaled to be in the interval $(0,1)$ by multiplying it by a constant \code{mrg31k3p\_NORM\_cl}, which is defined to be $1/2147483648$ for generating double-precision random numbers. If to generate random integers, \code{temp} is not scaled. At the end of generating random numbers, streams are transferred back to global memory through the function \fct{streamsFromPrivate}.
\begin{framed}
\begin{lstlisting}[language=C,basicstyle=\small,label={lst:uniformkernel},caption={Using streams to produce random U(0,1) numbers}]
__kernel void mrg31k3pMatrix(
  __global int* streams,
  __global float* out){

@@

Now we use the streams created in section \ref{createstreams} to generate two vectors of double-precision i.i.d. $U(0,1)$ random numbers with the function \fct{runif}. To view the generated random numbers, we need to convert them to R vectors or matrices, by doing this, the random numbers are moved from the global memory of the GPU device to the host.
\begin{knitrout}
\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
\begin{verbatim}
as.vector(clrng::runif(n = 6, streams = myStreamsGpu1, Nglobal = c(1, nrow(myStreamsGpu1)),
    type = "double"))
## [1] 0.7353245 0.6142074 0.1100781 0.6487742 0.3661944 0.1088229
as.vector(clrng::runif(n = 10, streams = myStreamsGpu2, Nglobal = c(1,
    nrow(myStreamsGpu1)), type = "double"))[-(1:6)]
## [1] 0.5330548 0.9783798 0.9151238 0.8509746
\end{verbatim}
\end{kframe}
\end{knitrout}
The arguments of the \fct{clrng::runif} are described as follows:
\begin{itemize}
\itemsep0em 
  \item \code{n}: a vector of length 2 specifying the row and column number if to create a matrix, or a number specifying the length if to create a vector.
  \item \code{streams}: streams for random number generation.
  \item \code{Nglobal}: global index space. Default is set as $(64,8)$.
  \item \code{type}: ``double'' or ``float'' or ``integer'' format of generated random numbers.
  \item \code{verbose}: verbose print extra information if >1.
\end{itemize}
\fct{runif} returns the same sequence of random numbers because it uses the same streams \code{myStreamsGpu1} and \code{myStreamsGpu2} here. Reusing \code{myStreamsGpu1} will produce a vector different from \code{sim_1}. Because each time a random number is generated, the current state of the stream advances by one position. %Below shows the \code{myStreamsGpu1} after it being used.

@@

Listing \ref{lst:normal} shows a fragment of the kernel that generates standard Gaussian random numbers. Aside from using local memory and the part that does the transformation, this kernel is the same with the kernel for generating uniform random numbers shown in Listing \ref{lst:uniformkernel}. If developers want to use other Gaussian transform methods, this kernel is easy to be adapted and incorporated in other \proglang{R} packages. The kernel executes over a 2-dimensional NDRange with $1 \times 2$ work-groups. \code{part[0]} and \code{part[1]} correspond to $R$ and $\Theta$ in the formulas respectively. As the work-items in a work-group proceed at differing rates, \code{barrier(CLK_LOCAL_MEM_FENCE)} ensures correct ordering of memory operations to local memory, so that Gaussian random numbers $(X_1,Y_1), \dots, (X_n,Y_n)$ are generated in pairs correctly, errors such as $(X_n, Y_{(n-1)})$ or $(X_{(n-1)}, Y_{(n)})$ are avoided.
%
\begin{framed}
\begin{lstlisting}[language=C,basicstyle=\small,label={lst:normal},caption={Using streams to produce standard normal random numbers},escapeinside={(*@}{@*)}]
__kernel void mrg31k3pMatrix(
  __global int* streams,
  __global float* out){

(*@{\hspace*{3cm}\vdots}@*)
local float part[2];
float sinOrCosPart1, addForSine = get_local_id(1) * PI;
if(get_local_id(1)){
  fact = TWOPI_mrg31k3p_NORM_cl;
}
streamsToPrivate(streams,g1,g2,startvalue);
for(DrowBlock = 0, Drow=get_global_id(0), DrowStart = Drow * NpadCol;
    DrowBlock < Nrow;
    Drow += get_global_size(0), 
    DrowBlock +=get_global_size(0), DrowStart += DrowStartInc){
    DrowInBounds = Drow < Nrow;
    for(DcolBlock = 0, Dcol=get_global_id(1), Dentry = DrowStart + Dcol;
        DcolBlock < Ncol;
        DcolBlock += get_global_size(1), Dentry += get_global_size(1)){

@@

We generate a large-size matrix of 100 million double-precision Gaussian random numbers, and compare the run-time between using \fct{stats::rnorm} and \fct{clrng::rnorm}. The best performance we have seen using \fct{clrng::rnorm} is around 120 times faster with $512 \times 128$ work-items than using the \fct{stats::rnorm}. The difference in elapsed time becomes larger when matrix size goes larger.
\begin{knitrout}
\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
\begin{verbatim}
streams <- createStreamsGpu(n = 512 * 128)
system.time(clrng::rnorm(c(10000,10000), streams=streams, 
                             Nglobal=c(512,128), type="double"))
##    user  system elapsed 
##   0.092   0.004   0.393
\end{verbatim}
\end{kframe}
\end{knitrout}
\begin{knitrout}
\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
\begin{verbatim}
system.time(matrix(stats::rnorm(10000^2),10000,10000))
##    user  system elapsed 
##   4.629   0.348   4.971
\end{verbatim}
\end{kframe}
\end{knitrout}

@@

\begin{knitrout}
\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
\begin{verbatim}
#using GPU
system.time(fisher.sim(month_gpu, 1e6, streams=streams, 
            type="double", returnStatistics=FALSE, Nglobal = c(256,64)))
##    user  system elapsed 
##   0.617   0.008   2.156
\end{verbatim}
\end{kframe}
\end{knitrout}
We requested one million simulations on GPU while the actual number of simulation is 1015808, and got 409524 cases whose test statitsic is below the observed threshold, so the p-value from \fct{clrng::fisher.sim} for this table is about 0.40443, which is close enough to the p-value from \fct{stats::fisher.test}: 0.40448. \fct{clrng::fisher.sim} takes about 0.69 seconds, \pkg{clrng} accounts for 4.64\% of the elapsed time on CPU.

@@

\begin{knitrout}
\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
\begin{verbatim}
#using GPU
system.time(clrng::fisher.sim(week_GPU, 1e7, streams=streams,
            type="double", returnStatistics=FALSE,Nglobal = c(256,64)))
##    user  system elapsed 
##   0.491   0.023  10.210
\end{verbatim}
\end{kframe}
\end{knitrout}
The ``week'' table has a much smaller p-value: around 0.0001249672, which should require a larger number of simulations to get a more accurate p-value. With more than ten million simulations, we get 1205 cases and a p-value around 0.000120472 by \fct{fisher.sim}. \fct{stats::fisher.test} takes 91.6 seconds, while \fct{fisher.sim} takes about 4.282 seconds, the elapsed time is decreased to about 4.67\% of the time taken on CPU.

@@

gpuBatchMatrix::multiplyLowerDiagonalBatch(
  simMat, maternCov, diagMat, zmatGpu,
  diagIsOne = 1L, # diagonal of L is one
  transformD = "sqrt", # take the square root of each element of D
  Nglobal,
  Nlocal,
  NlocalCache)
\end{verbatim}
\end{kframe}
\end{knitrout}
\begin{gather}
 \begin{bmatrix}  L_{1} \\ L_{2} \\L_{3} \\ \vdots
 \end{bmatrix} 
 *
  \begin{bmatrix}
   D_{1} \\ D_{2} \\D_{3} \\ \vdots
   \end{bmatrix} 
   *
   \begin{bmatrix}
   Z_{11} & Z_{12}
   \end{bmatrix}
  =
 \begin{bmatrix}
   L_1D_1Z_{11} & L_1D_1Z_{12} \\
   L_2D_2Z_{11} & L_2D_2Z_{12} \\
   L_3D_3Z_{11} & L_3D_3Z_{12} 
  \end{bmatrix}
\end{gather}
Step 7, Finally, plot the 10 simulated Gaussian random surfaces.
\begin{knitrout}
\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
\begin{verbatim}
library(raster)
simRaster = brick(myRaster, nl = ncol(simMat)*nrow(paramsGpu))
values(simRaster) = as.vector(as.matrix(simMat))
names(simRaster) = apply(expand.grid('par',1:nrow(paramsGpu), 
                                     'sim', 1:ncol(simMat)), 1, paste, collapse='')
\end{verbatim}
\end{kframe}
\end{knitrout}
\begin{figure}[H]
\centering
\begin{knitrout}
\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
\begin{verbatim}
par(mar=rep(0.1, 4))
for(D in names(simRaster)) {
  plot(extent(simRaster))
  plot(simRaster[[D]], legend=FALSE, add=TRUE)
}
\end{verbatim}
\end{kframe}
\includegraphics[width=0.45\textwidth]{figure/maternplot-1} 
\includegraphics[width=0.45\textwidth]{figure/maternplot-2} 
\includegraphics[width=0.45\textwidth]{figure/maternplot-3} 
\includegraphics[width=0.45\textwidth]{figure/maternplot-4} 
\includegraphics[width=0.45\textwidth]{figure/maternplot-5} 
\includegraphics[width=0.45\textwidth]{figure/maternplot-6} 
\includegraphics[width=0.45\textwidth]{figure/maternplot-7} 
\includegraphics[width=0.45\textwidth]{figure/maternplot-8} 
\includegraphics[width=0.45\textwidth]{figure/maternplot-9} 
\includegraphics[width=0.45\textwidth]{figure/maternplot-10} 
\end{knitrout}
\caption{Simulated Gaussian surface\label{fig:5} }
\end{figure}

>>>>

%THE PROBLEM:  each process needs to be independent, there's a risk of overlap.  solution: streams.  mrg31kp3 a stream is a vector of three positive integers.  initial state is first stream, compute new streams sequentially.
\subsection{Creating streams}\label{createstreams}
Random number generation is more complicated when the RNG algorithm is run in parallel processes in \proglang{R}, there is a risk that the generated sequences of random numbers have correlation. \pkg{clrng} uses multiple distinct streams that are used in work-items that executes in parallel on a GPU device. A popular way of obtaining multiple streams is to take an RNG with a long period and cut the RNG's sequence into very long disjoint pieces of equal length $Z$, and use each piece as a separate stream. Creating a new stream amounts to computing its starting point. In general, a stream object contains three elements: the current state of the stream, the initial state of the stream (or seed), and the initial state of the current substream (by default it is equal to the seed). Streams are created sequentially in the way that whenever the user creates a new stream, the software automatically jumps ahead by Z steps to find its initial state, and the three states in the stream object are set to it. Each of these streams can also be partitioned into substreams with equally-spaced starting points \citep{l2002object, rLEC15a}.

In \proglang{clRNG} library, the MRG31k3p RNG's entire period of length approximately $2^{185}$ is divided into approximately $2^{51}$ non-overlapping streams of length $Z = 2^{134}$. Each stream can be further partitioned into substreams of length $W = 2^{72}$, although this is not currently implemented in \pkg{clrng}. The state (and seed) of each stream is a vector of six 31-bit integers. This size of state is appropriate for having streams running in work-items on GPU cards, while providing a sufficient period length for most applications. The initial state of the first stream (also called ``initial seed for the package'' in \proglang{clRNG} and ``initial'' in \pkg{clrng}) for the MRG31k3p is by default $(12345, 12345, 12345, 12345, 12345, 12345)$.

%The function \fct{clrngMrg31k3pCreateStreams} from \proglang{clRNG} creates streams on the host using the MRG31k3p RNG. If we want to use the streams in work-items on a GPU device, the streams have to be copied from the host to the global memory of the GPU device, and then each work-item copies the current states only of the streams to its private memory. 
\pkg{clrng} is able to create streams both on the host and on the GPU device. The \fct{createStreamsCpu} function does the former. The \proglang{R} output below creates 4 streams on the host.
%is an interface function of \fct{clrngMrg31k3pCreateStreams} that creates streams on host as \proglang{R} matrices, see the \proglang{R} output below, where we created 4 streams on the host using \fct{createStreamsCpu}, after transposing \code{myStreamsCpu}, each column of the matrix is a stream, creating streams on host makes it easy for us to view and check the streams behavior. Substreams are not created in \pkg{clrng} because currently we use one stream per work-item. 
\begin{knitrout}
\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
\begin{verbatim}
# creating streams on CPU
library("gpuR")
library("clrng")
myStreamsCpu <- createStreamsCpu(n=4, initial=12345)
t(myStreamsCpu)
##               [,1]       [,2]       [,3]       [,4]
## current.g1.1 12345  336690377  502033783  739421137
## current.g1.2 12345  597094797 1322587635 1475938232
## current.g1.3 12345 1245771585 1964121530  730262207
## current.g2.1 12345   85196284 1949818481 1630192198
## current.g2.2 12345  523477687 1607232546  324551134
## current.g2.3 12345 2094976052 1462898381  795289868
## initial.g1.1 12345  336690377  502033783  739421137
## initial.g1.2 12345  597094797 1322587635 1475938232
## initial.g1.3 12345 1245771585 1964121530  730262207
## initial.g2.1 12345   85196284 1949818481 1630192198
## initial.g2.2 12345  523477687 1607232546  324551134
## initial.g2.3 12345 2094976052 1462898381  795289868
\end{verbatim}
\end{kframe}
\end{knitrout}
\fct{createStreamsCpu} has two arguments:
\begin{itemize}
\itemsep0em 
  \item \code{n}: number of streams to create.
  \item \code{initial}: vector of length 6, recycled if shorter.
\end{itemize}
We move streams to the GPU by converting them to a `\code{vclMatrix}'. %\fct{createStreamsGpu} is adapted from the clrngMrg31k3pCreateStreams() and creates streams directly on a GPU device. Setting a seed same as the MRG31k3p default seed in \fct{createStreamsGpu}, we produce on device exactly the same streams as those created on host as follows. Notice that the code is just for showing \fct{createStreamsGpu} and \fct{createStreamsCpu} are \proglang{R} interface functions for \fct{clrngMrg31k3pCreateStreams} on GPU and CPU respectively, in practice it is not recommended to create a new stream that has the same initial seed as another existing stream in one simulation program.
\begin{knitrout}
\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
\begin{verbatim}
# creating streams on GPU
myStreamsGpu = vclMatrix(myStreamsCpu)
myStreamsGpu2 = createStreamsGpu(n=4, initial=12345)
\end{verbatim}
\end{kframe}
\end{knitrout}
\fct{createStreamsGpu} creates streams directly on the GPU and returns a `\code{vclMatrix}', which makes it slightly more efficient when the number of streams is large. However, any data is lost when the \proglang{R} session is terminated.
%also has the two arguments \code{n} and \code{initial}. \fct{createStreamsGpu} is similar to \code{.Random.seed} in \proglang{R}. By setting the RNG status with the \proglang{R} function \fct{set.seed}, we are able to repeat the sequence of random seeds, which is an important quality criteria of RNGs. Similarly in \fct{createStreamsGpu} and \fct{createStreamsCpu}, with the \code{initial} argument, we make the sequence of streams repeatable.

\subsection{Using streams to generate uniform random numbers}
The streams created sequentially are then used by the GPU's work-items to generate random numbers. In \pkg{clrng} each work-item takes one distinct stream to generate random numbers, so the number of streams should always be greater than the number of total work-items in use. The main part of the kernel (\proglang{OpenCL} functions for execution on the device) for generating uniform random numbers is shown in Listing \ref{lst:uniformkernel}. Users can set \code{verbose>1} in \fct{runif} to print out the kernels. Kernels are written in the \proglang{OpenCL C} language, in which \code{__kernel} declares a function as a kernel, pointer kernel arguments must be declared with an address space qualifier, for example \code{__global} and \code{__local}. Here the pointers to \code{streams} and to the output matrix \code{out} on the global memory are passed to the kernel as arguments. \code{Nrow} and \code{Ncol} represent row and column number of the matrix \code{out} respectively. \code{NpadCol} is the internal number of columns of \code{out}, \code{NpadStreams} is the internal number of columns of \code{streams} (these variables are defined in macros not shown here). Each stream's current state is copied to the private memory of each work-item by the function \fct{streamsToPrivate}, where \code{g1} and \code{g2} point to the first three and second three elements of stream states. \code{index} represents the position of the work-item when mapping it to a one-dimensional data structure. The function \fct{clrngMrg31k3pNextState} generates an uniform random integer called \code{temp} between 1 and 2147483647,  the value of \code{temp} is then scaled to be in the interval $(0,1)$ by multiplying it by a constant \code{mrg31k3p\_NORM\_cl}, which is defined to be $1/2147483648$ for generating double-precision random numbers. If to generate random integers, \code{temp} is not scaled. At the end of generating random numbers, streams are transferred back to global memory through the function \fct{streamsFromPrivate}.
\begin{framed}
\begin{lstlisting}[language=C,basicstyle=\small,breaklines=true,label={lst:uniformkernel}]
__kernel void mrg31k3pMatrix(
  __global int* streams,
  __global float* out){

@@

Now we use the streams created in section \ref{createstreams} to generate two vectors of double-precision i.i.d. $U(0,1)$ random numbers with the function \fct{runif}. To view the generated random numbers, we need to convert them to R vectors or matrices, by doing this, the random numbers are moved from the global memory of the GPU device to the host.
\begin{knitrout}
\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
\begin{verbatim}
as.vector(clrng::runif(n = 6, streams = myStreamsGpu, Nglobal = c(1, nrow(myStreamsGpu)),
    type = "double"))
## [1] 0.7353245 0.6142074 0.1100781 0.6487742 0.3661944 0.1088229
\end{verbatim}
\end{kframe}
\end{knitrout}
The arguments of the \fct{clrng::runif} are described as follows:
\begin{itemize}
\itemsep0em 
  \item \code{n}: a vector of length 2 specifying the row and column number if to create a matrix, or a number specifying the length if to create a vector.
  \item \code{streams}: streams for random number generation.
  \item \code{Nglobal}: global index space. Default is set as $(64,8)$.
  \item \code{type}: ``double'' or ``float'' or ``integer'' format of generated random numbers.
  \item \code{verbose}: verbose print extra information if >1.
\end{itemize}
\fct{runif} returns the same sequence of random numbers because it uses the same streams \code{myStreamsGpu} and \code{myStreamsGpu2} here. Reusing \code{myStreamsGpu} will produce a vector different from \code{sim_1}. Because each time a random number is generated, the current state of the stream advances by one position. %Below shows the \code{myStreamsGpu} after it being used.

@@

Listing \ref{lst:normal} shows a fragment of the kernel that generates standard Gaussian random numbers. Aside from using local memory and the part that does the transformation, this kernel is the same with the kernel for generating uniform random numbers shown in Listing \ref{lst:uniformkernel}. If developers want to use other Gaussian transform methods, this kernel is easy to be adapted and incorporated in other \proglang{R} packages. The kernel executes over a 2-dimensional NDRange with $1 \times 2$ work-groups. \code{part[0]} and \code{part[1]} correspond to $R$ and $\Theta$ in the formulas respectively. As the work-items in a work-group proceed at differing rates, \code{barrier(CLK_LOCAL_MEM_FENCE)} ensures correct ordering of memory operations to local memory, so that Gaussian random numbers $(X_1,Y_1), \dots, (X_n,Y_n)$ are generated in pairs correctly, errors such as $(X_n, Y_{(n-1)})$ or $(X_{(n-1)}, Y_{(n)})$ are avoided.
%
\begin{framed}
\begin{lstlisting}[language=C,basicstyle=\small,label={lst:normal},breaklines=true, escapeinside={(*@}{@*)}]
__kernel void mrg31k3pMatrix(
  __global int* streams,
  __global float* out){

(*@{\hspace*{0cm}\vdots}@*)
local float part[2];
float sinOrCosPart1, addForSine = get_local_id(1) * PI;
if(get_local_id(1)){
  fact = TWOPI_mrg31k3p_NORM_cl;
}
streamsToPrivate(streams,g1,g2,startvalue);
for(DrowBlock = 0, Drow=get_global_id(0), DrowStart = Drow * NpadCol;
    DrowBlock < Nrow;
    Drow += get_global_size(0), 
    DrowBlock +=get_global_size(0), DrowStart += DrowStartInc){
    DrowInBounds = Drow < Nrow;
    for(DcolBlock = 0, Dcol=get_global_id(1), Dentry = DrowStart + Dcol;
        DcolBlock < Ncol;
        DcolBlock += get_global_size(1), Dentry += get_global_size(1)){

@@

We generate a large-size matrix of 100 million double-precision Gaussian random numbers, and compare the run-time between using \fct{stats::rnorm} and \fct{clrng::rnorm}. The best performance we have seen using \fct{clrng::rnorm} is around 120 times faster with $512 \times 128$ work-items than using the \fct{stats::rnorm}. The difference in elapsed time becomes larger when matrix size goes larger.
\begin{knitrout}
\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
\begin{verbatim}
streams <- createStreamsGpu(n = 512 * 128)
system.time(clrng::rnorm(c(10000,10000), streams=streams, 
                             Nglobal=c(512,128), type="double"))
##    user  system elapsed 
##   0.160   0.004   0.461
\end{verbatim}
\end{kframe}
\end{knitrout}
\begin{knitrout}
\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
\begin{verbatim}
system.time(matrix(stats::rnorm(10000^2),10000,10000))
##    user  system elapsed 
##   4.629   0.348   4.971
\end{verbatim}
\end{kframe}
\end{knitrout}

@@

\begin{knitrout}
\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
\begin{verbatim}
#using GPU
system.time(fisher.sim(month_gpu, 1e6, streams=streams, 
            type="double", returnStatistics=FALSE, Nglobal = c(256,64)))
##    user  system elapsed 
##   1.138   0.019   2.691
\end{verbatim}
\end{kframe}
\end{knitrout}
We requested one million simulations on GPU while the actual number of simulation is 1015808, and got 409524 cases whose test statitsic is below the observed threshold, so the p-value from \fct{clrng::fisher.sim} for this table is about 0.40443, which is close enough to the p-value from \fct{stats::fisher.test}: 0.40448. \fct{clrng::fisher.sim} takes about 0.69 seconds, \pkg{clrng} accounts for 4.64\% of the elapsed time on CPU.

@@

\begin{knitrout}
\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
\begin{verbatim}
#using GPU
system.time(clrng::fisher.sim(week_GPU, 1e7, streams=streams,
            type="double", returnStatistics=FALSE,Nglobal = c(256,64)))
##    user  system elapsed 
##   0.895   0.001  10.601
\end{verbatim}
\end{kframe}
\end{knitrout}
The ``week'' table has a much smaller p-value: around 0.0001249672, which should require a larger number of simulations to get a more accurate p-value. With more than ten million simulations, we get 1205 cases and a p-value around 0.000120472 by \fct{fisher.sim}. \fct{stats::fisher.test} takes 91.6 seconds, while \fct{fisher.sim} takes about 4.282 seconds, the elapsed time is decreased to about 4.67\% of the time taken on CPU.

@@

gpuBatchMatrix::multiplyLowerDiagonalBatch(
  simMat, maternCov, diagMat, zmatGpu,
  diagIsOne = 1L, # diagonal of L is one
  transformD = "sqrt", # take the square root of each element of D
  Nglobal,
  Nlocal,
  NlocalCache)
\end{verbatim}
\end{kframe}
\end{knitrout}
\begin{gather}
 \begin{bmatrix}  L_{1} \\ L_{2} \\L_{3} \\ \vdots
 \end{bmatrix} 
 *
  \begin{bmatrix}
   D_{1} \\ D_{2} \\D_{3} \\ \vdots
   \end{bmatrix} 
   *
   \begin{bmatrix}
   Z_{11} & Z_{12}
   \end{bmatrix}
  =
 \begin{bmatrix}
   L_1D_1Z_{11} & L_1D_1Z_{12} \\
   L_2D_2Z_{11} & L_2D_2Z_{12} \\
   L_3D_3Z_{11} & L_3D_3Z_{12} 
  \end{bmatrix}
\end{gather}
Step 7, Finally, plot the 10 realizations.
\begin{knitrout}
\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
\begin{verbatim}
library(raster)
simRaster = brick(myRaster, nl = ncol(simMat)*nrow(paramsGpu))
values(simRaster) = as.vector(as.matrix(simMat))
names(simRaster) = apply(expand.grid('par',1:nrow(paramsGpu), 
                                     'sim', 1:ncol(simMat)), 1, paste, collapse='')
\end{verbatim}
\end{kframe}
\end{knitrout}
\begin{figure}[H]
\centering
\begin{knitrout}
\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
\begin{verbatim}
par(mar=rep(0.1, 4))
for(D in names(simRaster)) {
  plot(extent(simRaster))
  plot(simRaster[[D]], legend=FALSE, add=TRUE)
}
\end{verbatim}
\end{kframe}
\includegraphics[width=0.45\textwidth]{figure/maternplot-1} 
\includegraphics[width=0.45\textwidth]{figure/maternplot-2} 
\includegraphics[width=0.45\textwidth]{figure/maternplot-3} 
\includegraphics[width=0.45\textwidth]{figure/maternplot-4} 
\includegraphics[width=0.45\textwidth]{figure/maternplot-5} 
\includegraphics[width=0.45\textwidth]{figure/maternplot-6} 
\includegraphics[width=0.45\textwidth]{figure/maternplot-7} 
\includegraphics[width=0.45\textwidth]{figure/maternplot-8} 
\includegraphics[width=0.45\textwidth]{figure/maternplot-9} 
\includegraphics[width=0.45\textwidth]{figure/maternplot-10} 
\end{knitrout}
\caption{Simulated Gaussian surface\label{fig:5} }
\end{figure}
